An investigation into using scagnostics to find interesting projections of high-dimensional data

### 2018-10-18 scagnostics.Rmd
Minor changes in the text.

Figure \ref{fig:bbhSimulation} shows a scatterplot matrix analogue to that shown in Figure \ref{fig:neutronStarSPLOM} but where we have removed a subset of less informative parameters to obtain a clear picture. Strong association can be seen between the parameters ra, dec and time, where the first two describe the position of the event in the sky, and time is the merging time (in GPS units). Moreover, the relation between dec and time is not functional, which motivates us to use the TIC index for this example (a drawback is however the run time, this data sample contains 10k points, making optimisation of the MINE index functions very inefficient). In general we observe non-standard distributions for all variables, as well as various patterns of association between sets of variables. These patterns are representative of the likelihood function, since most priors are flat, or build such to capture growth with volume rather than distance. This is still efficient thanks to special sampling techniques that allow jumps in parameter space.

### Data pre-pocessing
We follow the same steps of data pre-processing as described above for the first example. First, since constraints on the masses of two systems observed via gravitational waves are strongly correltated, we again remove the smaller mass m2 from the set of parameters. We then rescale all parameters to the range from 0 to 1.

@@

As a first test we use the full dataset and compare results obtained with different index functions. Here we expect to recover the relation between the correlated variables dec, time and ra. We compare the final view using three different metrics in Figure \ref{fig:bbhGuided}, showing the views with (local?) maximum of the TIC index (left), spines2D index (middle) and for comparison the maximum of 1-convex (right). Indeed the TIC and spline index identify dec, time and ra parameters and show views of the 2-d band that is embedded in three dimensions \textcolor{red}{maybe having tour of those three dimensions would be helpful here}. Notably, the exact views differ based on the index function selected, optimising TIC is showing us the band "from the side". On the other hand, the index based on the scagnostics member convex is too noisy to find any views of interest, and the final  view is essentially equal to the starting plane in this case. All index values for the three final views are listed in Table \ref{tab:indexTableBBH}, where we see that indeed the splines2D index does not change between the first two views, while the TIC index is increased for the first view wrt the second. Finally the value of 1-convex is clearly larger for the first view as compared to the final view, but the optimisation did not succeed for this noisy index function.

@@

We next investigate the dependence of the final result on the input starting plane. In particular if there are many local maxima the effect may be large, and many random starting planes may be needed to find a global maximum. To this end we first reduce the dataset dimensionality to include only two types of parameters: first, spin related parameters (i.e. alpha, theta\_jn, chi\_tot and chi\_p), and second position related parameters (i.e. ra, dec and distance). Note that this includes a number of variables for which we have established association and one dimensional structure above, i.e. ra and dec; chi\_tot and chi\_p; and theta\_jn. We generate a small set of random starting planes and optimise the splines2D index (it is prefered here over the TIC index as it is much more efficient to compute on large data sets). We compare six final views identified in this manner (i.e. differing only in randomly selected starting plane) in Fig. \ref{fig:bbhGuided3}. We note large variation in final views, only one of the six tries identified the view of ra vs dec, which however clearly has the maximum index value amongst all views. \textcolor{red}{add axes to plot, then we can make statement about how "far" the other views are from the "best" one.}

Finally, given the results above, it makes sense to also consider smaller parameter-subspaces to look for stronger association. (A procedure to first scout the parameter space for large index values is not efficent in this case because of the strong dependence on rotation angles in this paricular setup.) One subspace for which we found strong association between three variables is using the randomly selected parameter set (m1, ra, chi\_tot, alpha, distance, dec). Optimising again the splines2D index value allows us to find the final view shown in Figure \ref{constructedExample} (left) \textcolor{red}{(The small constribution of alpha is hidden in the axes here, middle plot is without rescaling, therefore same size contribution from dec an alpha to y axis)}, which we can reproduce from the original parameters as shown in the middle, as a combination of the position parameters ra and dec with the spin parameter alpha. For comparison we also show the dependence between ra and dec only, which is fuzzy compared to the ideal view identified by the PP optimisation. \textcolor{red}{This result is unexpected, it can be understood from ????? (Does it mean that knowing the precession angle gives more precise information about the location? Knowing ra uniquely determines dec-alpha rather than dec it seems.)}

>>>>

Figure \ref{fig:bbhSimulation} shows a scatterplot matrix analogue to that shown in Figure \ref{fig:neutronStarSPLOM} but where we have removed a subset of less informative parameters to obtain a clear picture. Strong association can be seen between the parameters ra, dec and time, where the first two describe the position of the event in the sky, and time is the merging time (in GPS units). Moreover, the relation between dec and time is not functional, which motivates us to use the TIC index for this example (a drawback is however the run time, this data sample contains 10k points, making optimisation of the MINE index functions very inefficient). In general we observe non-standard distributions for all variables, as well as various patterns of association between sets of variables.
\textcolor{red}{These patterns are representative of the likelihood function, since most priors are flat, or build such to capture growth with volume rather than distance. This is still efficient thanks to special sampling techniques that allow jumps in parameter space.}

### Data pre-pocessing
We follow the same steps of data pre-processing as described above for the first example. First, since constraints on the masses of two systems observed via gravitational waves are strongly correlated, we again remove the smaller mass m2 from the set of parameters. We then rescale all parameters to the range from 0 to 1.

@@

As a first test we use the full dataset and compare results obtained with different index functions. Here we expect to recover the relation between the correlated variables dec, time and ra. We compare the final view using three different metrics in Figure \ref{fig:bbhGuided}, showing the views with maximum of the TIC index (left), spines2D index (middle) and for comparison the local maximum found when optimising 1-convex (right). Indeed the TIC and splines2D index identify dec, time and ra parameters and show views of the 2-d band that is embedded in three dimensions \textcolor{red}{maybe having tour of those three dimensions would be helpful here}. Notably, the exact views differ based on the index function selected, optimising TIC is showing us the band "from the side". On the other hand, the index based on the scagnostics member convex is too noisy to find any views of interest, and the final  view is essentially equal to the starting plane in this case. All index values for the three final views are listed in Table \ref{tab:indexTableBBH}, where we see that the splines2D index does not change between the first two views, while the TIC index is increased for the first view compared the second. Finally the value of 1-convex is clearly larger for the first view as compared to the final view, but the optimisation did not succeed for this noisy index function.

@@

We next investigate the dependence of the final result on the input starting plane. In particular if there are many local maxima the effect may be large, and many random starting planes may be needed to find a global maximum. To this end we first reduce the dataset dimensionality to include only two types of parameters: first, spin related parameters (i.e. alpha, theta\_jn, chi\_tot and chi\_p), and second position related parameters (i.e. ra, dec and distance). Note that this includes a number of variables for which we have established association and one dimensional structure above, i.e. ra and dec; chi\_tot and chi\_p; and theta\_jn. We generate a small set of random starting planes and optimise the splines2D index (it is prefered here over the TIC index as it is much more efficient to compute on large data sets). We compare six final views identified in this manner (i.e. differing only in randomly selected starting plane) in Fig. \ref{fig:bbhGuided3}. We note large variation in final views, only one of the six tries identified the view of ra vs dec, which however clearly has the maximum index value amongst all views. \textcolor{red}{We see that all final fiews pick up some combination of parameters for which we expect structure, but the structure of the data distribution and corresponding index function dependence is such that one easily gets stuck in local maxima. One way out could be the scouting phase discussed above, however this requires some additional improvements, because non-rotation-invariance of the distribution and index function make this approach intractable here.}

Finally, given the results above, it makes sense to also consider smaller parameter-subspaces to look for new association. One subspace for which we found strong association between three variables is using the randomly selected parameter set (m1, ra, chi\_tot, alpha, distance, dec). Optimising again the splines2D index value allows us to find the final view shown in Figure \ref{constructedExample} (left) \textcolor{red}{(The small constribution of alpha is hidden in the axes here, middle plot is without rescaling, therefore same size contribution from dec an alpha to y axis)}, which we can reproduce from the original parameters as shown in the middle, as a combination of the position parameters ra and dec with the spin parameter alpha. For comparison we also show the dependence between ra and dec only, which is fuzzy compared to the ideal view identified by the PP optimisation. \textcolor{red}{This result is unexpected, it can be understood from ????? (Does it mean that knowing the precession angle gives more precise information about the location? Knowing ra uniquely determines dec-alpha rather than dec it seems.)}
